---
title: "RNA"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
library(MultiAssayExperiment)
library(here)
library(ggplot2)
library(knitr)
library(microbenchmark)
library(nipals)
```

```{r}
## utils
source(here("src/source-utils.R"))
```

# DNA

## Data

```{r}
gastru.mae <- readRDS(here('output/scnmtseq_gastrulation_mae_826-cells_orderedFeatures.rds'))  # MAE experiment
keep_cells <- (gastru.mae$lineage10x_2 != 'ExE_ectoderm') & !is.na(gastru.mae$lineage10x_2)
gastru.mae <- gastru.mae[, keep_cells,]  # MAE experiment
```


### NIPALS vs Weighted EMPCA runtimes

**Clarification**

The presentation showed the results for NIPLAS without weights while it was meant to convey that it is possible to do weighted analysis as well. The weighted algorithm is stochastic (and slow!) and can converge to a local minima at times.
We did see some improvements at times and this behaviour is something that needs to be further characterised.

```{r}
met_promoter <- t(assay(gastru.mae, 'met_promoter'))
wt_met_promoter <- t(assay(gastru.mae, 'wt_met_promoter'))
```


```{r}
microbenchmark_nipals <- function(value_mat, wt_mat, subset, times = 3L) {
  value_mat <- subset_pn(value_mat, n = subset[1], p = subset[2])
  wt_mat <- subset_pn(wt_mat, n = subset[1], p = subset[2])
  
  
  mb <- microbenchmark(
    nipals = nipals::nipals(value_mat, ncomp = 2, center = TRUE, scale = TRUE, fitted = FALSE, force.na = TRUE, gramschmidt = FALSE, verbose = FALSE, maxiter = 500, tol = 1e-8),
    empca = nipals::empca(value_mat, w = wt_mat, ncomp = 2, center = TRUE, scale = TRUE, fitted = FALSE, gramschmidt = FALSE, verbose = FALSE, maxiter = 500, tol = 1e-8),
    times = times, unit = 's'
  )
  
  mb_df <- summary(mb)[1:7] %>% data.frame(row.names = 1) 
  mb_df$algo = rownames(mb_df)
  mb_df$P <- subset[2]
  mb_df$N <- subset[1]
  mb_df
  
}

```

```{r}
## test
microbenchmark_nipals(value_mat = met_promoter, wt_mat = wt_met_promoter, subset = c(20, 40), times = 3L)
```

```{r}
mb_results <- list()
ns <- c(815)
ps <- c(100, 200, 400, 800)
mb_results <- list()

for (n in ns) {
  for (p in ps) {
    mb_results[[sprintf("N%sP%s",n,p)]] <- microbenchmark_nipals(value_mat = met_promoter, wt_mat = wt_met_promoter, subset = c(n, p), times = 3L)
    
  }
}
saveRDS(mb_results, file = 'savedata/mb_results.rds')
```


```{r}
apply_nipals_assay <- function(mae, assay_type, subset = c(1e5, 1000)) {
  dna_assay <- t(assay(mae, assay_type))
  dna_assay <- subset_pn(dna_assay, n = subset[1], p = subset[2])
  dna_assay_wt <- t(assay(mae, paste0('wt_',assay_type)))
  dna_assay_wt <- subset_pn(dna_assay_wt, n = subset[1], p = subset[2])
  
  nipals = nipals::nipals(dna_assay, ncomp = 2, center = TRUE, scale = TRUE, fitted = FALSE, force.na = TRUE, gramschmidt = FALSE, verbose = FALSE, maxiter = 500, tol = 1e-8)
    empca = nipals::empca(dna_assay, w = dna_assay_wt, ncomp = 2, center = TRUE, scale = TRUE, fitted = FALSE, gramschmidt = FALSE, verbose = FALSE, maxiter = 500, tol = 1e-8)
    return(list(nipals = nipals, empca = empca))
}

apply_nipals <- function(mae, assays, subset = c(1e5, 1000) ) {
  lapply(named_list(assays), function(z) {
    res <- apply_nipals_assay(mae = mae, assay_type = z, subset = subset)
    saveRDS(res, file = sprintf('savedata/%s-%s.rds', z, paste0(subset, collapse = '-')))
    res
  })
}

nipals_assays <- c("met_genebody", "met_promoter", "acc_p300", 
"acc_CTCF", "acc_DHS", "acc_promoter")
nipals_all <- apply_nipals(mae = gastru.mae, assays = nipals_assays, subset = c(1e4, 1000))
saveRDS(nipals_all, file = 'savedata/nipals-all.rds')
```

```{r}
mb_results <- readRDS('savedata/mb_results.rds')
```

```{r}
nipals_all <- readRDS('savedata/nipals-all.rds')
```

```{r}
mb_results <- rbindListWithNames(mb_results)
```

```{r}
ggplot(mb_results , aes(x = P, y = mean, col = algo)) +
    geom_errorbar(aes(ymin=lq, ymax=uq), width=0.1) +
    geom_line() +
    geom_point()
```

```{r}
assay_type <- 'met_promoter'

pca_res <- nipals_all[[assay_type]]
col.values = c("#ADD8E6", "#7390C7", "#3948A9", "#00008B")

ggplot_pcs <- function(scores, col, legend.title, col.values = NULL, ...) {
  scores <- data.frame(scores)
  scores <- scores[rownames(colData(gastru.mae)),]
  scores$col <- col
  p <- ggplot(scores , aes(x = PC1, y = PC2, col = col)) + 
    geom_point() + theme_bw() +
    guides(col = guide_legend(title = legend.title)) +
    labs(...)
    
    if (!is.null(col.values)) {
      p <- p + scale_color_manual(values = col.values)
    }
    p
}
```

```{r}
set.seed(42)
nstage <- length(unique(gastru.mae$stage))
nlineage <- length(unique(gastru.mae$lineage10x_2))
ntotal <- nstage + nlineage
col_pallete <- sample(gg_color_hue(n=ntotal), size = ntotal)

stage_cols <- col_pallete[seq_len(nstage)]
lineage_cols <- col_pallete[-seq_len(nstage)]
```



```{r}
p1 <- ggplot_pcs(scores = nipals_all$met_promoter$nipals$scores, col = gastru.mae$stage, legend.title = 'Stage', col.values = stage_cols)
p2 <- ggplot_pcs(scores = nipals_all$met_promoter$empca$scores, col = gastru.mae$stage, legend.title = 'Stage', col.values = stage_cols)
p3 <- ggplot_pcs(scores = nipals_all$met_promoter$nipals$scores, col = gastru.mae$lineage10x_2, legend.title = 'Lineage', col.values = lineage_cols)
p4 <- ggplot_pcs(scores = nipals_all$met_promoter$empca$scores, col = gastru.mae$lineage10x_2, legend.title = 'Lineage', col.values = lineage_cols)
library(egg)
ggarrange(p1, p2, p3, p4)
```

```{r}
coldata <- data.frame(colData(gastru.mae))
nipals_scores <- lapply(named_list(names(nipals_all)), function(x) {
  df <- data.frame(nipals_all[[x]]$nipals$scores)
  df <- df[rownames(coldata),]
  df <- cbind(df, coldata)
  df
})

nipals_scores <- rbindListWithNames(nipals_scores, new_col = 'assay')
                                   
empca_scores <- lapply(named_list(names(nipals_all)), function(x) {
  df <- data.frame(nipals_all[[x]]$empca$scores)
  df <- df[rownames(coldata),]
  df <- cbind(df, coldata)
  df
})

empca_scores <- rbindListWithNames(empca_scores, new_col = 'assay')
```

```{r}
# ggplot(empca_scores, aes(PC1, PC2)) + geom_point(aes(col = stage), alpha = 0.7) + facet_wrap(.~assay)
# ggplot(nipals_scores, aes(PC1, PC2)) + geom_point(aes(col = lineage10x_2), alpha = 0.7) + facet_wrap(.~assay)
ggplot(nipals_scores, aes(PC1, PC2)) + geom_point(aes(col = stage), alpha = 0.7) + facet_wrap(.~assay)  + theme_classic() +
  theme(strip.text.x = element_text(size = 10, face = 'bold'))
ggsave(filename = 'figures/nipals-assays.pdf', width = 10, height = 7)
```






