---
title: "Compare SCTransform and scran PCA clusters using given phenotypes"
author: "Al J Abadi"
params:
  on_hpc: !r Sys.info()['user'] != 'alabadi'
  on_mac: !r Sys.info()['user'] == 'alabadi'
  rerun: false
  cache: true
output:
  html_document:
    toc: true
---

```{r, warning=FALSE, message=FALSE}
library(SingleCellExperiment)
library(MultiAssayExperiment)
library(here)
library(ggplot2)
library(mixOmics)
```


```{r, echo=FALSE}
knitr::opts_chunk$set(eval = TRUE, cache = params$cache)
source(here("src/source-utils.R"))
```

## Data

```{r}
sce_matching <- readRDS(here("output/sce-matching-1pct-cells-expressing.rds"))
gastru.mae <- readRDS(here('output/scnmtseq_gastrulation_mae_826-cells_orderedFeatures.rds'))  # MAE experiment
```

## Compare normalisation methods

**Calculate PCA-based silhouette widths:**

Function to perform PCA on normalied counts and find an optimum number of components that maximises the (distribution of) silhouette width:

```{r}
pca_normalised_counts <- function(norm_counts, group=gastru.mae$stage_lineage, ncomp="auto", comps_try = 10) {
  
  calc_opt_ncomp <- FALSE
  if ( ncomp == "auto") {
    calc_opt_ncomp <- TRUE
    ncomp <- comps_try
  }
  pca_res <- mixOmics::pca(t(norm_counts), ncomp = ncomp)
  
  out <- list(pca_res = pca_res)
  library(cluster)
  
  if (!calc_opt_ncomp) {
    sils <- silhouette(x = as.integer(factor(group)), dist = dist(pca_res$variates$X[,ncomp]))[,"sil_width"]
  } else{
    
    comp_opt <- 1
    sils <- silhouette(x = as.integer(factor(group)), dist = dist(pca_res$variates$X[,comp_opt]))[,"sil_width"]
    for (i in seq_len(ncomp)[-1]) {
      sils_new <- silhouette(x = as.integer(factor(group)), dist = dist(pca_res$variates$X[,seq_len(i)]))[,"sil_width"]
      
      p_val <- t.test(sils_new, sils, alternative = "greater")$p.value
      
      if (p_val < 0.05) {
        comp_opt <- i
        sils <- sils_new
      } else {
        break()
      }
    }
    out$ncomp_maxsilhouette = comp_opt
  }
  out$silhouette <- sils
  
  return(out)
}
```

function to plot mean-variance:

```{r}
plot_meanvar <- function(norm_count_mat, col = "#D37B0A", ...) {
  ggplot(data.frame(mean = rowMeans(norm_count_mat), variance = rowVars(norm_count_mat))) +
    geom_point(aes(x = mean, y = variance), col = col, alpha = 0.75) + theme_bw() +
    labs(...)
}
```


### DESeq2

```{r, eval=params$rerun}
library(DESeq2)
deseq2_vst <- DESeq2::varianceStabilizingTransformation(counts(sce_matching))
saveRDS(deseq2_vst, file = here("output/deseq2-vst.rds"))
```

```{r, eval=!params$rerun}
deseq2_vst <- readRDS(here("output/deseq2-vst.rds"))
```

#### mean-variance
```{r}
plot_meanvar(deseq2_vst)
```


```{r, eval=params$rerun}
deseq2_clusters <- pca_normalised_counts(norm_counts = deseq2_vst,  group = gastru.mae$stage_lineage, ncomp = "auto")
save(deseq2_clusters, file = here("output/deseq2_clusters.RData"))
```
```{r, eval=!params$rerun}
load(here("output/deseq2_clusters.RData"))
```

#### silhouette

```{r}
## optimum ncomp
deseq2_clusters$ncomp_maxsilhouette
## median sil. width
median(deseq2_clusters$silhouette)
```

### scran

```{r, eval=params$rerun}
scran_clusters <- pca_normalised_counts(norm_counts = logcounts(sce_matching),  group = gastru.mae$stage_lineage, ncomp = "auto")
save(scran_clusters, file = here("output/scran_clusters.RData"))
```
```{r, eval=!params$rerun}
load(here("output/scran_clusters.RData"))
```

#### mean-variance

```{r}
plot_meanvar(logcounts(sce_matching))
```

#### silhouette

```{r}
## optimum ncomp
scran_clusters$ncomp_maxsilhouette
## median sil. width
median(scran_clusters$silhouette)
```

### SCTransform

#### UMI corrected logcounts

```{r, eval=params$rerun}
library(sctransform)
counts_vst <- vst(as(counts(sce_matching), "dgCMatrix"),n_genes = NULL, min_cells = 0, return_cell_attr = TRUE, return_corrected_umi = TRUE)
saveRDS(counts_vst, file = here("output/sctrandorm-vst-out-allgenes.rds"))
```

```{r, eval=!params$rerun}
counts_vst <- readRDS(here("output/sctrandorm-vst-out-allgenes.rds"))
```

Get UMI corrected logcounts so that the effects of cell UMI library size is regressed out in gene variance calculation:
```{r}
counts_umi_crrected <- as(counts_vst$umi_corrected, "matrix")
## logcounts of UMI corrected data
logcounts_umi_corrected <- log(counts_umi_crrected+1)
```

##### PCA

```{r, eval=params$rerun}
sct_clusters <- pca_normalised_counts(norm_counts = logcounts_umi_corrected,  group = gastru.mae$stage_lineage, ncomp = "auto")
save(sct_clusters , file = here("output/sct_clusters.RData"))
```
```{r, eval=!params$rerun}
load(here("output/sct_clusters.RData"))
```

##### silhouette

```{r}
sct_clusters$ncomp_maxsilhouette
```

```{r}
sil_title <- function(values, caption) {
  sprintf("%s: median = %s  -  mean = %s", caption, round(median(values), 3), round(mean(values), 3))
}
hist_values(sct_clusters$silhouette, fill_col = 19, x = "silhouette width", title = sil_title(values = sct_clusters$silhouette, caption = "SCTransform - UMI Corrected"))
hist_values(scran_clusters$silhouette, fill_col = 19, x = "silhouette width", title =  sil_title(values = scran_clusters$silhouette, caption = "Scran - logcounts"))
```

##### mean-variance

```{r}
plot_meanvar(logcounts_umi_corrected)
```


#### Pearson residuals

##### PCA

```{r, eval=params$rerun}
sct_clusters_pearson_residuals <- pca_normalised_counts(norm_counts = counts_vst$y,  group = gastru.mae$stage_lineage, ncomp = "auto")
save(sct_clusters_pearson_residuals , file = here("output/sct_clusters_pearson_residuals.RData"))
```
```{r, eval=!params$rerun}
load(here("output/sct_clusters_pearson_residuals.RData"))
```

#### silhouette

```{r}
sct_clusters_pearson_residuals$ncomp_maxsilhouette
```
```{r}
hist_values(sct_clusters_pearson_residuals$silhouette, fill_col = 19, x = "silhouette width", title = sil_title(values = sct_clusters_pearson_residuals$silhouette, caption = "SCTransform - Pearson residuals"))
hist_values(scran_clusters$silhouette, fill_col = 19, x = "silhouette width", title =  sil_title(values = scran_clusters$silhouette, caption = "Scran - logcounts"))
```

#### mean-variance

```{r}
plot_meanvar(counts_vst$y)
```

## sample plots

### sctranform - Pearson residuals
```{r}
plotIndiv(sct_clusters_pearson_residuals$pca_res, pch= 16, group = gastru.mae$stage_lineage, comp = c(1,2))
plotIndiv(sct_clusters_pearson_residuals$pca_res, pch= 16, group = gastru.mae$stage_lineage, comp = c(1,3))
plotIndiv(sct_clusters_pearson_residuals$pca_res, pch= 16, group = gastru.mae$stage_lineage, comp = c(3,4))
```


### scran
```{r}
plotIndiv(scran_clusters$pca_res, pch= 16, group = gastru.mae$stage_lineage, comp = c(1,2))
plotIndiv(scran_clusters$pca_res, pch= 16, group = gastru.mae$stage_lineage, comp = c(1,3))
plotIndiv(scran_clusters$pca_res, pch= 16, group = gastru.mae$stage_lineage, comp = c(3,4))
```

## Results

* scran-, umi-corrected- and pearson residual-normalised counts all yielded the best cluster quality (measured using silhouette width) using 4 componenets.
* DESeq2 performs best in stablising variance, while for Pearson residuals the dependency is pronounced.
* Pearson residuals yields a better median silhouette width for `stage_lineage`-based clusters (0.142) compared to other methods ( ~ 0).


## add Pearson residuals to SCE

```{r, eval=params$rerun}
assay(sce_matching, "pearson_residuals") <- counts_vst$y
assay(sce_matching, "DESeq2") <- deseq2_vst
saveRDS(sce_matching, file = here("output/sce-matching-1pct-cells-expressing_all-normalistion-methods.rds"))
```
