---
title: "Filter and order RNA data based on feature variance"
author: "Al J Abadi"
params:
  on_hpc: !r Sys.info()['user'] != 'alabadi'
  on_mac: !r Sys.info()['user'] == 'alabadi'
output:
  html_document:
    toc: true
---

```{r, warning=FALSE, message=FALSE}
library(SingleCellExperiment)
library(here)
library(ggplot2)
library(scater)
```


```{r, echo=FALSE}
knitr::opts_chunk$set(eval = TRUE, cache = FALSE)
```

## rna data

The `rna` assay is simply the `logcounts(sce)` so no filtering so far. We need to filter near-zero-variance genes:

### read the matching SCE

Read the SCE with matching cells (copied from __Hackathon/data folder):

```{r}
sce_matching <- readRDS(here("data/gastrulation/rna/parsed/SingleCellExperiment_matching-cells.rds"))
dim(sce_matching)
```

### histpgram of % of cells not expressing every given gene

```{r}
hist(rowData(sce_matching)$pct_dropout_by_counts, main = "Histogram of percent of cell dropouts for genes")
```

### mean expression histogram

Look at mean expression histogram for all genes:
```{r}
hist_mean_exp <- function(sce, fill_col = "#3381A7") {
  rowdata <- as.data.frame(rowData(sce))
  ggplot(rowdata) + geom_histogram(aes(log10_mean_counts), bins = 50, fill = fill_col) + theme_bw()
}

hist_mean_exp(sce_matching)
```

### table of gene dropout rates

```{r}
table_dropouts <- function(sce, pct_cells_expresing = c(0, 1, 5, 10)) {
  out <- data.frame(pct_cells_expressing = pct_cells_expresing, Number_of_genes = NA)
  out$Number_of_genes <- sapply(pct_cells_expresing, function(x){
    sum(rowData(sce)$pct_dropout_by_counts <= 100-x)
  })
  out
}

table_dropouts(sce_matching)

```


### genes to be expressed in at least 1% of cells

```{r}
sce_matching <- sce_matching[rowData(sce_matching)$pct_dropout_by_counts <= 99 ,]
dim(sce_matching)
```

### mean expression histogram after filtering

```{r}
hist_mean_exp(sce_matching)
```

A great number of genes with low mean expression are potentially affected by drop-out.

## predict NA lineages

```{r, eval=FALSE}
rna_data <- t(gastru.mae[["rna"]])
missing_lineages <- is.na(colData(gastru.mae)$lineage10x_2)
known_lineages <- colData(gastru.mae)$lineage10x_2[!missing_lineages]
## run pls on known lineages
plsda_rna <- mixOmics::plsda(X = rna_data[!missing_lineage,], Y = known_lineages)
## find the distance that performs best
plsda_perf <- perf(plsda_rna, validation = "Mfold", folds = 8) ## centroids
plsda_perf$error.rate
# $overall
#        max.dist centroids.dist mahalanobis.dist
# comp1 0.5115431      0.6026731        0.6026731
# comp2 0.3134872      0.2916160        0.2612394
# 
# $BER
#        max.dist centroids.dist mahalanobis.dist
# comp1 0.7503743      0.5296016        0.5296016
# comp2 0.6308841      0.2322845        0.2853030
predicted_lineages <- predict(plsda_rna, newdata = rna_data[missing_lineages,], dist = 'centroids.dist')
predicted_lineages$class
```


### Output

Save the output expression data:

```{r, eval=FALSE}
saveRDS(sce_matching, file = here("output/sce-matching-1pct-cells-expressing.rds"))
```


```{r, eval=FALSE}
knitr::purl("filter-rna-data.Rmd")
```

